Writing Trash Bin

### 2024-03-15 paper.qmd
finished taxonomy section

A large drive of the field of uncertainty visualisation, has been the idea that "uncertainty visualisation" (defined as one word) is somehow different to "uncertainty" "visualisation" (separately defined as two words). When "uncertainty visualisation" is defined as a single word we lose some nuance in our understanding of the driving force behind some issues, and problems with visualisation and problems with uncertainty become conflated. This appears in the literature as a mountain of seemingly conflicting statements about what is, or is not, an "uncertainty visualisation". For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in most "uncertainty visualisation" experiments [@Ibrekk1987] (*Cite Gap 10*). @Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of mass and would already be considered an uncertainty visualisation were our statistic of interest the population mean.

These conflicting ideas and lacking definition of what is means for a visualisation to be an "uncertainty visualisation" is a cornerstone of some clearly questionable results within this field.

## Taxonomies of uncertainty visualisation

Similar to "uncertainty" exactly what defines an "uncertainty visualisation" is not clear. Often, the visualisaitons are simply categorised according to features, however exactly what *is* or *is not* an uncertainty visualisation is rarely explained.

There are some taxonomies of visual uncertainty. These taxonomies seem to seek to define "uncertainty visualisation" and a combined term rather than as two separate words. We largely disagree with this conceptualization of uncertainty visualisations as it blurs the line between the mathematical estimations and assumptions (the uncertainty part) and the visual depiction of those mathematical objects (the visualisation part). While we will not focus on these taxonomies, it is still important to mention them and highlight the reasons they are not good practice.

There are a small handful of taxonomies (or typologies) that are specifically for uncertainty visualisations. @Kinkeldey2014 identified give categories of uncertainty based on the work of other literature reviews of uncertainty visualisation. The five categories are: (1) explicit/implicit (directly mapping or showing multiple outcomes) (2) intrinsic/extrinsic (using existing symbols e.g. colour value, or new objects e.g. grids) (3) visually integral/separable (can or cannot be separated from the data) (4) coincidence/adjacent (if data and uncertainty are in integrated or separate views (5) static/dynamic (animation/interaction). Of the groups they identified, they actually only used (4), (2), and (5), left out (1) because most visualisations are explicit, and (3) corresponds to (1) in most cases [@Kinkeldey2014].

@Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor). However, because the term "PDF", a statistical object that describes a random variable that is typically a one dimensional function, is used to describe both the data and the uncertainty for all dimensions, it is hard to understand how this dimensionality should work.

@Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space that was defined by the axis "domain expertise" and "continum of discreteness" that scaled from "point estimate" to "continuous distribtion".

@Thomson2005 titled their paper "A typology for visualizing uncertainty" however it is a taxonomy for the specific types of uncertainty that we may want to visualise and a way to calculate that uncertainty. While this paper is more about mathematics than visualisation, it is clearly written as a precursor to visualisation.

![Example of different uncertainty visualisation taxonomies](paperscreenshots/collage2.jpg){#fig-location}

These taxonomies are useful in the sense that they often identify areas where there is an over or under supply of uncertainty visualisations. Additionally taxonomies do highlight a concept of uncertainty dimensionality that was not directly covered in the uncertainty taxonomy, however the distinction between uncertainty as a single value versus as a PDF could be considered a sub-consideration of uncertainty that's level is statistical so it can be quantified. While uncertainty visualisation taxonomies are not the standard practice (generating a taxonomy of uncertainty and then developing methods to depict that uncertainty is) [@Kinkeldey2014], it is still important to acknowledge the existence of "uncertainty visualisation typologies" and why they may miss the mark in what is needed from the uncertainty visualisaiton literature.

The first issue with these taxonomies is that most of these taxonomies are created by observing existing uncertainty visualisations. This means under researched gaps in the taxonomy could be due to a lack of need just as well as a lack of attention, and the benefit of trying to create visualisations that land in uncharted areas of the taxonomy are unclear.

The second issue is that there does not seem to be anything these taxonomies offer that would not be better established by separate taxonomies for uncertainty and visualisation. These taxonomies seem to depict uncertainty visualisation as a field within itself, rather than a specific case of information visualisation that requires more attention to mathematics than usual. The difference between "uncertainty visualisations" and "data visualisations" is not technically in the visual element, it is mathematical. @Kinkeldey2014 almost acknowledges this in their own paper that discusses an uncertainty visualsiation taxonomy when they claim "future typologies should take different categories of tasks into account (1) communication tasks (2) analytical tasks (3) exploratory task", a common typology for information visualisation in general. The process of understanding and estimating uncertainty requires knowledge of the data, the statistical methods used to make an estimate, and the assumptions of a model. Visualising the statistics that represent uncertainty should be no different than depicting the statistics that represent any other element of a graphic, and therefore there seems to be no reason to have a separate taxonomy for uncertainty visualisation.

## Contextual Information
> Edit Note: This section should probably be moved to the visualisation specific section, although that depends on the tone that I end up with after the rewrites of the earlier sections.

Some uncertainty work, specifically uncertainty visualisation, focus on the context surrounding the uncertainty. These are things like spatial data, temporal data, category data, etc. While other discussions of uncertainty may consider this distinction more relevant to their work, here we will consider this contextual information. Contextual information is important for interpretation and understanding of graphics, but it does not in of itself generate uncertainty. This is not unique to the properties of spatial information, other statistics information that is important for modelling and assumptions, such as non-linearity, class levels, etc do not contribute to uncertainty. Operations on data, inherrent randomness, and assumptions that create bias generate uncertainty, which is what we are talking about here. A confidence interval is a confidence interval whether it came from spatial data, temporal, or cross sectional data.

>>>>

## Establishing misunderstood concepts
Before discussing the issues facing uncertainty visualisation due to a poor concept *of* uncertainty visualisation, we want to establish two ideas that should be considered in conjunction with the lacking definition of uncertainty. The first is the sidestepping of the "uncertainty" definition problem by suggesting "uncertainty visualisation" is inherently different. Some authors may consider "uncertainty" to be poorly defined, but "uncertainty visualisation" to be well established. We will spend the next section explaining why that is not the case by discussing several taxonomies of uncertainty visualisation. The second is a conflation of the contextual information with the visual information. This problem is not, in isolation an issue, however it does seem to be a *key* contributor to the misunderstanding that uncertainty is an attribute of data, and not a by product of inference. Both of these issues need to be understood in order for us to fully understand the issues surrounding uncertainty visualisation.

### Definitions of "uncertainty visualisation"
A large drive of the field of uncertainty visualisation, has been the idea that "uncertainty visualisation" (defined as one word) is somehow different to "uncertainty" "visualisation" (separately defined as two words). When "uncertainty visualisation" is defined as a single word we lose some nuance in our understanding of the driving force behind some issues, and problems with visualisation and problems with uncertainty become conflated. This appears in the literature as a mountain of seemingly conflicting statements about what is, or is not, an "uncertainty visualisation". For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in most "uncertainty visualisation" experiments [@Ibrekk1987] (*Cite Gap 10*). @Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of mass and would already be considered an uncertainty visualisation were our statistic of interest the population mean. These conflicting ideas and lacking definition of what is means for a visualisation to be an "uncertainty visualisation" is a cornerstone of some clearly questionable results within this field.

Since the concept of "uncertainty visualisation" is harder to define than "uncertainty", "uncertainty visualisation" taxonomies are typically less thorough than taxonomies for "uncertainty". They also tend to blur the line between the mathematical elements of uncertainty and the visual depiction of those mathematical objects. While we will not focus on these taxonomies, it is still important to mention them and highlight what we can learn from their success' and failures..

Some taxonomies of uncertainty highlight how confusing most authors find the concept, and how difficult it is to articulate the rules of an uncertainty visualisation when uncertainty itself if not defined. @Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).  However, because the term "PDF", a statistical object that describes a random variable that is typically a one dimensional function, is used to describe both the data and the uncertainty for all dimensions, it is hard to understand how this should work. @Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space defined by the "domain expertise" and "continuum of discreteness" (that scaled from "point estimate" to "continuous distribution"). This conceptualisation could be considered a visual extension of the law of large numbers, because as a point estimate is a single sample, as the sample size gets larger the distribution will appear to be a continuous. However, assumes every point estimate is a sample from the same distribution (a mean is not technically a sample from a prediction interval, but it is a sample from the sampling distribution) and it conflates discreteness from a too small a number of observations, discreteness from a precision problem, discreteness because that is the true nature of the variable, and discreteness as a visualisation choice. @Griethe2006 organised uncertainty visualisations into two cases (1) a hypothesis test was preformed to confirm the validity of the visualisation and (2) the visualisation has uncertainty depicted. This conceptualization conflates hypothesis testing and confidence intervals, and while they are related concepts, a hypothesis test is not necessarily uncertainty. Of course the authors of these papers did not intend for the taxonomy to be a complete description of uncertainty visualisations, however these distinctions that are often ignored are subtly important. (*Cite Gap 10*)

There are a small handful of taxonomies that highlight some interesting features in the way uncertainty is visualised. @Kinkeldey2014 categorised uncertainty according to whether or not it was:
1. Implicit (a sample) or explicit (a depiction of mass)
2. Intrinsic (alter existing symbols symbols to represent uncertainty) or extrinsic (add new objects to represent uncertainty)
3. visually integral or separable (the uncertainty can be separated from the data and read independently) 
4. coincidence (uncertainty and data are represented in the same plot) or adjacent
5. static or dynamic (animated, interactive ect)
Of the groups they identified, they actually only used (4), (2), and (5), left out (1) because most visualisations are explicit, and (3) corresponds to (1) in most cases [@Kinkeldey2014]. This taxonomy suggests interesting considerations for a visualisation of a distribution (even though it was not intended to be a visualisation of a distribution). A similar version of this taxonomy was presented by @uncertchap2022 who commented that visualisation can be organised into two categories, "graphical annotations of distributional properties" and "visual encodings of uncertainty" which functionally aline with the intrinsic/extrinsic distinction. The first four categories boil down to the question "should uncertainty and signal be conveyed together as one variable or separately as two". This consideration will become more relevant later in this review.

![This visualisation is here as a placeholder. I am going to replace it with an R coded visualisation that shows the 5 considerations for showing uncertainty in a plot according to @Kinkeldey2014 ](paperscreenshots/collage2.jpg){#fig-location}

Most of these taxonomies are created by observing existing uncertainty visualisations, which means they suffer from many of the same conceptual issues that the field has with uncertainty in general. Additionally, there does not seem to be anything these taxonomies offer that would not be better established by separate taxonomies for uncertainty and visualisation. The difference between "uncertainty visualisations" and "data visualisations" is not technically in the visual element, it is mathematical. @Kinkeldey2014 almost acknowledges this in their own paper that discusses an uncertainty visualsiation taxonomy when they claim "future typologies should take different categories of tasks into account (1) communication tasks (2) analytical tasks (3) exploratory task", a common typology for information visualisation in general. The process of understanding and estimating uncertainty requires knowledge of the data, the statistical methods used to make an estimate, and the assumptions of a model. Visualising the statistics that represent uncertainty should be no different than depicting the statistics that represent any other element of a graphic, this is something we will show in the following sections of this paper.

### Contextual Information
Some uncertainty work, specifically uncertainty visualisation, focus on the context surrounding the uncertainty, such as the spatial or temporal nature of the estimate. While other discussions of uncertainty may consider this distinction more relevant to their work, here we will consider this contextual information. Contextual information is important for interpretation and understanding of graphics, but it does not in of itself generate uncertainty. Its relevance to the field of "uncertainty visualisation" is through the percieved lack of available channels. For example, a choropleth map has already used colour to represent an estimate, and the concept of space to provide contextual information, there is little remaining room tany spatial

This is not unique to the properties of spatial information, other statistics information that is important for modelling and assumptions, such as non-linearity, class levels, etc do not contribute to uncertainty. Operations on data, inherrent randomness, and assumptions that create bias generate uncertainty, which is what we are talking about here. A confidence interval is a confidence interval whether it came from spatial data, temporal, or cross sectional data.

### 2024-03-15 writingdumpster.qmd
finished taxonomy section

Cite Extra: INCORRECT INTERPRETATION OF STUDIES. People will use the best available information to them. If you ask for a maximum, and give them a confidence interval... they will give you the maximum number that they are aware is still plausible. Instead of being shocked the participants were wrong, it is important to ask "what was required for them to be RIGHT".
  - HAVE SOME PERSPECTIVE. If you are paying people below minimum wage to do your experiment, they don't care. If a problem requires extensive mental labour to solve, they will liekly not do it.
  - This result being reported as "people interpret a confidence interval as a maximum and minimum despite being told it was a confidence interval" is simply not true. It is a false equivalence. 
  - all these studies show is if you give people the wrong information... they will be wrong. Shocking.

@@

>>>>

Cite Gap 10: Uncertainty Visualisation Taxonomies

# Extra Section Cite Gaps
Cite Gap - New Section: Logical fallacy interpretation of mathematically unanswerable questions. If I tell you my cousins age, and then ask you to guess my age, it would not be unreasonable to guess the age I provided for my cousin. Citing this work to state "people incorrectly believe my cousins age is my age" would
- Deterministic questions about a random event cannot be answered. 
- @uncertchap2022 took the results of other paper to mean "when participants viewed mean temperature forecasts that included 95% confidence intervals depicted as bars with end caps, they incorrectly believed that the error bars represented high and low temperatures" (this one will depend on that study)

@@

@Thomson2005 titled their paper "A typology for visualizing uncertainty" however it is a taxonomy for the specific types of uncertainty that we may want to visualise and a way to calculate that uncertainty. While this paper is more about mathematics than visualisation, it is clearly written as a precursor to visualisation.
